---
title: "Annoying things to do with `ggplot2`"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align ='center', warning = FALSE, message = FALSE,
                      fig.height = 4, fig.width = 5)
```

After learning the [basic syntax](file:///home/emanuela/io/website/emanuelafurfaro/annoying-ggplot.html), there is a number of things that are bit tricky (and often annoying!) to do with `ggplot2`. By "tricky" I just mean a number of things to customise which can make your plots look much nicer, but are easy to forget! This post is a collection of ggplot customisation that I often do and that I often forget how to do, hoping it can serve as a reference to others as well as to myself.

All throughout the post I will work with the well known `iris` dataset:

```{r}
require(ggplot2)
require(dplyr)
data(iris)
```


## Add title

There are different ways to add a title in ggplot2. One way is that of using `labs()` function and specify the title there:

```{r}
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15) +
  labs(title = "My title")
```


## Change colours

One of the first things I wanted to change in ggplot was the default colours. Suppose for example we want to colour the histogram of `Petal.Length` according to the species, just like we did above:

```{r, fig.show = "hide"}
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15) 
```

There are different ways to do it. Here we see two ways of doing it, one using the color brewer palettes given by `R` and one way of doing it manually. To see more details on this, among the many websites that discuss this topic, I like [this one](http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually).

First, note that we are working on the `fill` aesthetics, we will refer to the `scale_fill_*` functions. If the colour we want to customise is given in the `col` aesthetics or in th `alpha` aesthetics for example, we will use the `scale_colour_*` and `scale_alpha_*` functions respectively.

### Use RColorBrewer palettes

If we want to use a colour palette as provided by the `RColorBrewer` package, the `ggplot` function that allows us to it is:

```{r}
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  scale_fill_brewer(palette = "Greens")
```

Given the factor order, it will assign the lightest colour to the first one and the darkest to the last one. If we want to change it, we may just use the `direction` parameter:

```{r}
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  scale_fill_brewer(palette = "Greens", direction = -1)
```

Palettes of gray can be used with the function `scale_fill_gray()`.

### Manually change colours

In order to manually change colours we will use `scale_fill_manual()` function. There are different ways to specify colours in `R`, such as using their name, their number, their rgb or their hex code.


```{r}
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet"))
```


## Change the default order of categories

If you do not like the default order of the categories (here it is the alphadetical order so setosa, versicolor and virginica), you need to order the factor the way you like and then use ggplot on it. Alternatively you can specify levels in the ggplot function itself:

```{r, fig.show = "hide"}
ggplot(data = iris, aes(x = Petal.Length, fill = factor(Species, levels = c("virginica", "versicolor", "setosa")))) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet"))
```


## Modify axis

To modify axis title, we use the `labs()` function, while in order to modify axis ticks, we use specific arguments in the `theme()` function:

```{r, fig.width=3, fig.height=3, fig.show='hold', fig.align='center', out.extra='style="float:left"'}
# Change axis title
ggplot(data = iris, aes(y = Petal.Length, x = 0)) +
  geom_boxplot(alpha = 0.7, fill = "hotpink") +
  labs(x = "", y = "Petal length")

# Hide axis text
ggplot(data = iris, aes(y = Petal.Length, x = 0)) +
  geom_boxplot(alpha = 0.7, fill = "hotpink") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

# Hide axis ticks
ggplot(data = iris, aes(y = Petal.Length, x = 0)) +
  geom_boxplot(alpha = 0.7, fill = "hotpink") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank())
```


## Customise legend

You can change the legend name, move it into a different place (for example at the bottom) or hide it:

```{r, fig.width=3, fig.height=3, fig.show='hold', fig.align='center', out.extra='style="float:left"'}
# Change legend name
ggplot(data = iris, aes(y = log(Petal.Length), x = Species, fill = Species)) +
  geom_boxplot(alpha = 0.7)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet"), name = "") 

# Change legend position
ggplot(data = iris, aes(y = log(Petal.Length), x = Species, fill = Species)) +
  geom_boxplot(alpha = 0.7)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(legend.position = "bottom")

# Hide legend
ggplot(data = iris, aes(y = log(Petal.Length), x = Species, fill = Species)) +
  geom_boxplot(alpha = 0.7)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(legend.position = "")
```


## Customise font in legend, axis and title

All these customisations can easily be done with the `theme()` function, which allows to customize the non-data components of plots. Let us start with legend and title:

```{r, fig.width=3, fig.height=3, fig.show='hold', fig.align='center', out.extra='style="float:left"'}

# Change legend's title font and size
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(legend.title = element_text(size = 20, face = "italic", colour = "red"))

# Change legend's categories font and size
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(legend.text = element_text(size = 20, face = "italic", colour = "red"))

# Change title's font and size
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  labs(title = "Histogram of petal length") + 
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(plot.title = element_text(size = 10, face = "bold", colour = "red"))

```

Similarly, axis names and labels can be customised:


```{r, fig.width=3, fig.height=3, fig.show='hold', fig.align='center', out.extra='style="float:left"'}

# Change axis's title font and size
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  labs(title = "Histogram of petal length") + 
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(axis.title = element_text(size = 10, face = "bold", colour = "red")) 
# use axis.title.x and axis.title.y to change axis separately

# Change all titles at once
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  labs(title = "Histogram of petal length") + 
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(title = element_text(size = 10, face = "bold", colour = "red"))


# Rotate axis's labels, change font and size
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(col = "black", alpha = 0.7, bins = 15)  +
  labs(title = "Histogram of petal length") + 
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) +
  theme(axis.text = element_text(size = 10, face = "bold", colour = "red", angle = 45))
# use axis.text.x and axis.text.y to change axis separately
```


## Barplots

Here are some issues I often encounter when I work with barplots.

### Barplot with given bar length

Often we have a table with proportions or with aggregate data already on it. Hence we do not want the barplot to count and aggrgate the data but we just want it to plot as it is. In order to get this, we need to adjust the `stat` parameter. As default it is set to `stat = "count"`, but what we want in this case is actually "identity":

```{r}
iris_aggr <- iris %>%
  group_by(Species) %>%
  summarise(prop = round(n()/nrow(iris), 3))

ggplot(iris_aggr, aes(x = Species, y = prop, fill = Species)) +
  geom_bar(stat = "identity", alpha = 0.7)
```



### Pie chart

The pie chart is not one of the available functions in `ggplot`. Hence in order to get one we need to work a bit on the plot. It is basically a barplot, but we need to "reshape" it. First we build a barplot and then we will tweak it to make it circular:

```{r}
# Use the width = 1 to make it occupy the whole plot
bplot <- ggplot(iris_aggr, aes(x = "", y = prop, fill = Species)) +
  geom_bar(stat = "identity", width = 1)  +
  scale_fill_manual(values = c("hotpink", "violet", "darkviolet")) 

# Reshape coordinates to make it round and get rid of axis ticks and labels
bplot +  
  coord_polar("y", start=0) +
  labs(x = "", y = "") +
  theme(axis.text.x=element_blank())
```




<P ALIGN=RIGHT> <font size="3"> _This article was written by Emanuela Furfaro._</font> </P>