---
title: "Joining tables using dplyr: case study"
author: "Emanuela"
date: "May 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
require(tidyverse)

chefmozcuisine <- read_delim("data/chefmozcuisine.csv", delim = ",", col_names = T)
userprofile <- read_delim("data/userprofile.csv", delim = ",", col_names = T)
rating_final <- read_delim("data/rating_final.csv", delim = ",", col_names = T)
```

Let us explore the strcture of the three tables we just imported:

```{r}
str(userprofile)
str(chefmozcuisine)
str(rating_final)
```

And now focus on the three variables of rating contained in the `rating_final` table:

```{r}
rating_final %>% distinct(rating)
rating_final %>% distinct(food_rating)
rating_final %>% distinct(service_rating)
```


First we need to manipulate the data a bit. For example we want to make sure that IDs are character instead of numeric so that we do not accidentally calculate anything on them:

```{r}
rating_final <- rating_final %>% mutate(placeID = as.character(placeID))
chefmozcuisine <- chefmozcuisine %>% mutate(placeID = as.character(placeID))
```

Another frequent issue is missing data. In these tabes missingness is coded as "?" which is not the missingness symbol in `R`. Hence with a small additional manipulation:

```{r}
userprofile <- userprofile %>% na_if("?")
```

Now we are ready to combine all the tables and use them to answer questions. 

Which restaurant has got the best reviews?

```{r}
rating_final %>% 
  group_by(placeID) %>% 
  count(rating) %>% mutate(prop = prop.table(n))
```

Which restaurant has got the best reviews in terms of average rating?

```{r}
rating_final %>% 
  group_by(placeID) %>%
  summarise_at(vars(rating, food_rating, service_rating), 
               funs(mean, n())) %>%
  arrange(desc(rating_n), desc(rating_mean), desc(food_rating_mean), desc(service_rating_mean)) %>%
  left_join(chefmozcuisine) 
```

What is the favourite restaurant for smokers?

```{r}
userprofile%>% 
  filter(smoker == "true") %>%
  left_join(rating_final) %>%
  group_by(placeID) %>%
  summarise_at(vars(rating_mean = rating, food_rating_mean = food_rating, 
                    service_rating_mean = service_rating), 
               funs(mean, n())) %>%
  arrange(desc(rating_mean), desc(food_rating_mean), desc(service_rating_mean)) %>%
  left_join(chefmozcuisine) 

```

What is the favourite restaurant for people with a low budget?

```{r}
userprofile %>% 
  filter(budget == "low") %>%
  left_join(rating_final) %>%
  group_by(placeID) %>%
  summarise_at(vars(rating_mean = rating, food_rating_mean = food_rating, 
                    service_rating_mean = service_rating), 
               funs(mean, n())) %>%
  arrange(desc(rating_mean), desc(food_rating_mean), desc(service_rating_mean)) %>%
  left_join(chefmozcuisine) 
```
